---
title: Simulations To Optimize Subsampling Chinook Tissues for Run Timing Genetic Analysis
author: Thomas Buehrens (tbuehrens@dfw.wa.gov) & Jeremy Wilson (jeremy.wilson@dfw.wa.gov)
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"https://privatelands.wdfw.wa.gov/wdfwlogo_clrnotxt.png"\" style=\"float: right;width: 150px;\"/>')
   });
</script>

***

Last Updated `r format(Sys.time(), '%m/%d/%Y')`.

***

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r functions}
#==========================================
#function to inverse logit transform
#==========================================
ilogit<-function(x){exp(x)/(1+exp(x))}
#==========================================
#function to install or load packages
#==========================================
install_or_load_pack <- function(pack){
  create.pkg <- pack[!(pack %in% installed.packages()[, "Package"])]
  if (length(create.pkg))
    install.packages(create.pkg, dependencies = TRUE)
  sapply(pack, require, character.only = TRUE)
}
#==========================================
#function to install or load packages
#==========================================
make_run_list<-function(dat,min_samps){
  run_list<-c()
  for(i in 1:max(dat$week)){
    samps<-sample(dat$sample_number[dat$week==i],min(min_samps,length(dat$sample_number[dat$week==i])),replace = F)
    run_list<-c(run_list,samps)
  }
  return(run_list)
}
```


```{r packages }
packages<-c(
  "lubridate",
  "tidyverse",
  "LDATS",
  "ggplot2",
  "mgcv",
  "modelr"
)
install_or_load_pack(packages)
```


## User inputs
```{r user inputs, echo=FALSE}
#get date list to simulate proportions for
date_list<-read_csv("2022_Kalama_Chinook_DNA_Samples.csv")%>%
  mutate(Date=mdy(Date))%>%
  dplyr::select(date=Date)%>%
  arrange(date)%>%
  pull()
#how steep the change over from spring to fall is...larger value equals steeper slope and more dominant proportions on tails of season
slope<-2.0
#how much random walk noise do we want to simulate relative to linear change in proportions?
process_sd<-0.1

#formula for spline multinomial model
formula<-
  list(
    run2 ~ s(day,bs='ps',m=2,k =floor(length(run_list)/8-1)),
         ~ s(day,bs='ps',m=2,k =floor(length(run_list)/8-1))
    )
# formula<-
#   list(
#     run2 ~ s(day),
#          ~ s(day)
#     )
```

## Simulate "True" Chinook Genotype Proportions
```{r pressure, echo=FALSE}
#simulate some multinomial data:
days<-length(min(date_list):max(date_list))
logit<-tibble(spring= 2*slope + cumsum(rnorm(days,0,process_sd)) - ((4*slope)/days)*(1:days),
              summer=rep(0,days),
              fall= -2*slope  + cumsum(rnorm(days,0,process_sd)) + ((4*slope)/days)*(1:days),
              day=1:days,
              date=c(min(date_list) + 0:(days-1))
              )

logit_plt<-logit%>%
  pivot_longer(values_to = "logit",names_to = "run", cols= - c(day,date))

probs<-data.frame(t(apply(logit%>%dplyr::select(-c(day,date)),1,softmax)))%>%
  tibble()%>%
  mutate(day=1:days,date=c(min(date_list) + 0:(days-1)))%>%
  pivot_longer(values_to = "probs",names_to = "run", cols= - c(day,date))
  
ggplot(logit_plt,aes(x=date,y=logit,color=run))+
  geom_line()+
  ylab("logit(simulated proportions)")

ggplot(probs,aes(x=date,y=probs,color=run))+
  geom_line()+
  ylab("simulated proportions")
```

## Simulate Run Type for All Existing Samples
```{r}
sample_dat<-read_csv("2022_Kalama_Chinook_DNA_Samples.csv")%>%
  mutate(Date=mdy(Date))%>%
  dplyr::select(date=Date,sample_number=DNA_Sample_Code,)%>%
  mutate(week=week(date),week=week-(min(week)-1))%>%
  left_join(probs)%>%
  mutate(run=factor(run))%>%
  group_by(sample_number)%>%
  mutate(count=as.vector(rmultinom(n=1,size=1,prob=probs)))%>%
  mutate(run2=as.numeric(run)-1)%>%
  filter(count>0)
```

## Estimate "True" Proportions Using Subsample

min(weekly sample size, 3)
```{r}
weekly_ss<-3
run_list<-make_run_list(sample_dat,min_samps = weekly_ss)
#print(run_list)
remain_samples<-sample_dat%>%
  filter(!sample_number %in% run_list)

tdat<-sample_dat%>%
  filter(sample_number %in% run_list)

# this works (no spline)
mod1 <- gam(
  data = tdat,
  formula = formula,
  family=multinom(K=2)
  )


preds<-data.frame(predict(mod1,newdata=tibble(expand.grid(day=1:days)),type = "response"))%>%
  tibble()%>%
  setNames(c("fall","spring","summer"))%>%
  mutate(day=1:days,date=c(min(date_list) + 0:(days-1)),type="prediction")%>%
  pivot_longer(cols=c("fall","spring","summer"),values_to = "probs",names_to = "run")%>%
  bind_rows(probs%>%
              mutate(type="true")
            )

ggplot(preds,aes(x=date,y=probs,color=run,linetype=type))+
  geom_line()+
  labs(y="run type proportions",title=paste0("Total Sample Size = ",dim(tdat)[1]," , weekly sample = min(",weekly_ss,", weekly sample size)"))
```


min(weekly sample size, 4)
```{r}
add_ss<-1
weekly_ss<-weekly_ss + add_ss
run_list<-c(run_list,make_run_list(remain_samples,min_samps = add_ss))
#print(run_list)
remain_samples<-sample_dat%>%
  filter(!sample_number %in% run_list)


tdat<-sample_dat%>%
  filter(sample_number %in% run_list)

# this works (no spline)
mod1 <- gam(
  data = tdat,
  formula = formula,
  family=multinom(K=2)
  )



preds<-data.frame(predict(mod1,newdata=tibble(expand.grid(day=1:days)),type = "response"))%>%
  tibble()%>%
  setNames(c("fall","spring","summer"))%>%
  mutate(day=1:days,date=c(min(date_list) + 0:(days-1)),type="prediction")%>%
  pivot_longer(cols=c("fall","spring","summer"),values_to = "probs",names_to = "run")%>%
  bind_rows(probs%>%
              mutate(type="true")
            )

ggplot(preds,aes(x=date,y=probs,color=run,linetype=type))+
  geom_line()+
  labs(y="run type proportions",title=paste0("Total Sample Size = ",dim(tdat)[1]," , weekly sample = min(",weekly_ss,", weekly sample size)"))
```

min(weekly sample size, 5)
```{r}
add_ss<-1
weekly_ss<-weekly_ss + add_ss
run_list<-c(run_list,make_run_list(remain_samples,min_samps = add_ss))
#print(run_list)
remain_samples<-sample_dat%>%
  filter(!sample_number %in% run_list)


tdat<-sample_dat%>%
  filter(sample_number %in% run_list)

# this works (no spline)
mod1 <- gam(
  data = tdat,
  formula = formula,
  family=multinom(K=2)
  )



preds<-data.frame(predict(mod1,newdata=tibble(expand.grid(day=1:days)),type = "response"))%>%
  tibble()%>%
  setNames(c("fall","spring","summer"))%>%
  mutate(day=1:days,date=c(min(date_list) + 0:(days-1)),type="prediction")%>%
  pivot_longer(cols=c("fall","spring","summer"),values_to = "probs",names_to = "run")%>%
  bind_rows(probs%>%
              mutate(type="true")
            )

ggplot(preds,aes(x=date,y=probs,color=run,linetype=type))+
  geom_line()+
  labs(y="run type proportions",title=paste0("Total Sample Size = ",dim(tdat)[1]," , weekly sample = min(",weekly_ss,", weekly sample size)"))
```


min(weekly sample size, 10)
```{r}
add_ss<-5
weekly_ss<-weekly_ss + add_ss
run_list<-c(run_list,make_run_list(remain_samples,min_samps = add_ss))
#print(run_list)
remain_samples<-sample_dat%>%
  filter(!sample_number %in% run_list)


tdat<-sample_dat%>%
  filter(sample_number %in% run_list)

# this works (no spline)
mod1 <- gam(
  data = tdat,
  formula = formula,
  family=multinom(K=2)
  )



preds<-data.frame(predict(mod1,newdata=tibble(expand.grid(day=1:days)),type = "response"))%>%
  tibble()%>%
  setNames(c("fall","spring","summer"))%>%
  mutate(day=1:days,date=c(min(date_list) + 0:(days-1)),type="prediction")%>%
  pivot_longer(cols=c("fall","spring","summer"),values_to = "probs",names_to = "run")%>%
  bind_rows(probs%>%
              mutate(type="true")
            )

ggplot(preds,aes(x=date,y=probs,color=run,linetype=type))+
  geom_line()+
  labs(y="run type proportions",title=paste0("Total Sample Size = ",dim(tdat)[1]," , weekly sample = min(",weekly_ss,", weekly sample size)"))
```

all samples
```{r}
add_ss<-max(sample_dat%>%group_by(week)%>%summarise(n=n())%>%pull()) - weekly_ss
weekly_ss<-weekly_ss + add_ss
run_list<-c(run_list,make_run_list(remain_samples,min_samps = add_ss))
#print(run_list)
remain_samples<-sample_dat%>%
  filter(!sample_number %in% run_list)


tdat<-sample_dat%>%
  filter(sample_number %in% run_list)

# this works (no spline)
mod1 <- gam(
  data = tdat,
  formula = formula,
  family=multinom(K=2)
  )


preds<-data.frame(predict(mod1,newdata=tibble(expand.grid(day=1:days)),type = "response"))%>%
  tibble()%>%
  setNames(c("fall","spring","summer"))%>%
  mutate(day=1:days,date=c(min(date_list) + 0:(days-1)),type="prediction")%>%
  pivot_longer(cols=c("fall","spring","summer"),values_to = "probs",names_to = "run")%>%
  bind_rows(probs%>%
              mutate(type="true")
            )

ggplot(preds,aes(x=date,y=probs,color=run,linetype=type))+
  geom_line()+
  labs(y="run type proportions",title=paste0("Total Sample Size = ",dim(tdat)[1]," , weekly sample = min(",weekly_ss,", weekly sample size)"))
```


## Generate Final Run List
```{r make run list}
dat<-read_csv("2022_Kalama_Chinook_DNA_Samples.csv")%>%
  mutate(Date=mdy(Date))%>%
  dplyr::rename(sample_number=DNA_Sample_Code,date=Date)%>%
  mutate(week=week(date),week=week-(min(week)-1))

run_list<-tibble(sample_number=make_run_list(dat,min_samps=5))%>%
  mutate(in_sample=1)%>%
  full_join(dat,by="sample_number")

write_csv(run_list,file="run_list.csv")  
```

